A Dataset of Scratch Programs:\\Scraped, Shaped and Scored

### 2017-02-20 MSR-Scratch.tex
Small fixes

A number of works in the computing education research field attempt to assess the programming skills that novice programmers develop in the Scratch environment.
Some utilize program data collected during specific programming courses (e.g.,~\cite{meerbaum-salant_learning_2010, wilson_evaluation_2012, Maloney_2008}), while others utilize the dataset made available by the Lifelong Kindergarten Group at the MIT Media Lab~\cite{2017_scratch_dataset}, which contains data for Scratch projects created until 2012 (e.g.,~\cite{fields_2014, yang_2015, Dasgupta_2016}).
In addition to identifying indications of learning of programming concepts, static analysis of Scratch programs has also been performed for identifying code smells and bad programming practices within Scratch programs~\cite{Meerbaum_habits_2011, Aivaloglou_2016}, and automated quality assessment tools have been proposed (e.g., Hairball~\cite{boe_hairball:_2013} and Dr. Scratch~\cite{moreno_automatic_2014}).

While Scratch is receiving increasing interest as an introductory programming language, there is no recent dataset of Scratch programs available to the research community.
The one made available from the MIT Media Lab concerns projects created using the previous, initial version of the Scratch application, before the introduction of the Scratch 2 web programming interface in 2013.
It is since then that the popularity of Scratch started to increase\footnote{Monthly activity trends can be found at \url{https://scratch.mit.edu/statistics/}}.

@@

\subsection{Data collection}
To collect the data from the web interface of the Scratch project repository, we built a scraping program.
The web scraping program, called Kragle, starts by reading the Scratch projects page\footnote{\label{scratchpublic}\url{https://scratch.mit.edu/explore/projects/all/}} and thus obtains the project identifiers of projects that were most recently shared.
Subsequently, Kragle retrieves a JSON file for each of the listed projects\footnote{For a given project id $x$, the program's JSON representation can be obtained via \url{https://cdn.projects.scratch.mit.edu/internalapi/project/x/get}}.

We ran Kragle on March 2nd 2016 for 24 hours and, during that time, it obtained the JSON files for \nPrograms~projects. Out of those, we failed to parse and further analyze 2,367 projects due to technical difficulties with the JSON files.
Kragle, all scraped projects and our analysis files are available\footnotemark[\ref{dataseturl}].

Once we obtained the Scratch projects, we parsed the JSON files according to the specification of the format\footnote{\url{http://wiki.scratch.mit.edu/wiki/Scratch_File_Format_(2.0)}}.
This resulted in a list of used blocks per project, within the sprites and the stage of the project.
We also cross referenced all blocks with the Scratch wiki to determine the shapes and the category of all blocks.
For example, \texttt{When Green Flag Clicked} is a \emph{Hat block} from the \emph{Events category}.
We included blocks from Scratch extensions, such as the LEGO WeDo extensions, that were found in the dataset.

\subsection{Calculation of programming mastery scores}
%\fenia{Description of how the mastery metrics for       
%	[Abstraction]
%	,[Parallelism]
%	,[Logic]
%	,[Synchronization]
%	,[FlowControl]
%	,[UserInteractivity]
%	,[DataRepresentation]
%	,[Mastery]
%	,[Clones]
%	,[CustomBlocks]
%	,[InstancesSprites]
%	were calculated}
The projects were analyzed with Dr. Scratch, a tool that statically inspects Scratch projects source code to assign scores on seven dimensions of computational thinking: abstraction and problem decomposition, logical thinking, synchronization, parallelism, algorithmic notions of flow control, user interactivity and data representation. These dimensions are given a value from 0 to 3. The mastery score is the aggregated punctuation of the seven dimensions, and therefore ranges from 0 to 21.

@@

@@

\begin{table}[]
	\centering
	\begin{tabular}{llp{5.4cm}}
		\textbf{Table}& \textbf{Key} & \textbf{Attribute(Description)}\\
		\hline
		\textbf{\texttt{project}} & PK & \texttt{p\_id} (Scratch project ID)\\
		&  & 							\texttt{project-name} (name given to project)\\
		&  & 							\texttt{username} (author's Scratch 
		username)\\
		&  & 							\texttt{total-views} (project views number)\\
		&  & 							\texttt{total-remixes} (project remixes number)\\
		&  & 							\texttt{total-favorites} (total users favoriting)\\
		&  & 							\texttt{total-loves} (users `loving' the project)\\
		& & 							\texttt{is-remix} (calculated column, true if project is a remix of another one)\\
		\hline
		\textbf{\texttt{script}} & PK & \texttt{script\_id} (auto increment)\\
		& FK & 							\texttt{project\_id} (\texttt{\scriptsize{*\ldots1 project:p\_id}})\\
		& & 							\texttt{sprite-type} ([$sprite \vert stage \vert procDef$])\\
		& & 							\texttt{sprite-name} (name sprite the script is in)\\
		& & 							\texttt{script-rank} (project-level ranking of script)\\
		& & 							\texttt{coordinates} (x-y location of script in Scratch editor)\\
		& & 							\texttt{total-blocks} (calculated column, number of blocks comprising the script)\\
		\hline
		\textbf{\texttt{procedure}} & PK,FK & \texttt{script\_id} (\texttt{\scriptsize{1\ldots1 script:script\_id}})\\
		& & 							\texttt{proc-name} (name given to custom block)\\
		& & 							\texttt{total-args} (number of input arguments)\\
		\hline
		\textbf{\texttt{block}} & PK,FK  & \texttt{script\_id} (\texttt{\scriptsize{*\ldots1 script:script\_id}})\\
		& PK & \texttt{block-rank} (script-level ranking of block)\\
		& FK & \texttt{block-type} (\texttt{\scriptsize{*\ldots1 blockType:b-type}})\\
		&  & \texttt{parameter-1} (value of 1st input parameter)\\
		&  & ... \\
		&  & \texttt{parameter-24} (value 24th input parameter)\\
		\hline
		\textbf{\texttt{blockType}} & PK & \texttt{b-type} (Scratch name of predefined block)\\
		&  & \texttt{category} (Scratch block category)\\
		&  & \texttt{shape} (One of 8 block shapes in Scratch)\\
		&  & \texttt{is-input} (true if block receives user input)\\
		\hline
		\textbf{\texttt{remix}} & PK  & \texttt{remix\_id} (\texttt{\scriptsize{project\_id}} of created project)\\
		&  & \texttt{from-project\_id} (\texttt{\scriptsize{project\_id}} of original project)\\
		\hline
		\textbf{\texttt{grades}} & PK,FK & \texttt{project\_id} (\texttt{\scriptsize{1\ldots1 project:p\_id}})\\
% Original
%		& & \texttt{Abstraction} (Score for abstraction and problem decomposition)\\
%		& & \texttt{Parallelism} (Score for parallelism)\\
%		& & \texttt{Logic} (Score for logical thinking)\\
%		& & \texttt{Synchronization} (Score for synchronization)\\
%		& & \texttt{FlowControl} (Score for algorithmic notions of flow control)\\
%		& & \texttt{UserInteractivity} (Score for user interactivity)\\
%		& & \texttt{DataRepresentation} (Score for data representation)\\

		& & \texttt{Abstraction} (Score)\\
		& & \texttt{Parallelism} (Score)\\
		& & \texttt{Logic} (Score)\\
		& & \texttt{Synchronization} (Score)\\
		& & \texttt{FlowControl} (Score)\\
		& & \texttt{UserInteractivity} (Score)\\
		& & \texttt{DataRepresentation} (Score)\\
		& & \texttt{Mastery} (Dr. Scratch total mastery score)\\
		& & \texttt{Clones} (\# software clones)\\
		& & \texttt{CustomBlocks} (\# user defined blocks)\\
		& & \texttt{InstancesSprites} (True if project uses Scratch clones)\\
		\hline
	\end{tabular}
	\caption{Database schema: Tables and Attributes}
	\label{tbl-dbschema}
\end{table}

@@

The data of a project are stored in a relational database. Each \textbf{\texttt{project}} is identified by its Scratch project ID, stored in field \texttt{p\_id}, while its author is identified by the \texttt{username}\footnote{\label{fn-authorpage}Each project and author page can be accessed in the Scratch web interface at \url{https://scratch.mit.edu/projects/<p_id>} and \url{https://scratch.mit.edu/users/<username>} respectively}.
If a project is a remix of another one, the original project can be found in the \textbf{\texttt{remix}} table.
Table~\textbf{\texttt{grades}} stores the Dr. Scratch results for the programming mastery metrics per project.

@@

The schema of the database, outlined in Table~\ref{tbl-dbschema}, reflects the structure of the Scratch programs.
Projects contain sprites, which are entities with their own associated code.
The code is organized into \textbf{\texttt{script}}s, i.e., groups of Scratch code blocks, each script belonging a sprite named \texttt{sprite-name}.
In the example in Figure~\ref{fig-scratchExample} there are 3 scripts, one initiated when the green flag is clicked, one then the sprite is clicked, and the `backflip' custom block definition.
Those custom blocks are the equivalent of procedure definitions, with their names and arguments stored in the \textbf{\texttt{procedure}} table.

The coding elements in the Scratch environment are the \textbf{\texttt{block}}s, which can receive input parameters.
In the example in the figure, block `go to x:0 y:0' receives as input two numerical parameters.
Custom clock invocations are stored in this table just like regular blocks.
To cater for the varying number of input parameters that those blocks can have, this table can store up to the maximum number of input parameters found in the dataset, which is 24.
The ordering of the blocks in the scripts can be retrieved using the \texttt{block-rank} field, which represents a depth-first ranking of the blocks.
In the script of the `backflip' custom block definition in the example, the ranking of the blocks would be `define backflip', `set counter to', `+', `counter', followed by the rest of the blocks.
The clocks can be of different shapes and categories.
In table \textbf{\texttt{blockType}} we have stored the encoding of all types of Scratch blocks and the blocks from Scratch extensions.

@@

\begin{table}[]
	\centering
	\begin{tabular}{lr}
		\hline
		Projects & 250,163 \\
		Non-empty projects & 233,491 \\
		Projects analyzed by Dr. Scratch & 231,050 \\
		Remixes of other projects & 12,167 \\
		Usernames & 109,960 \\
		Scripts & 4,049,356 \\
		Blocks & 36,085,654 \\
		Custom block definitions & 209,769 \\
		\hline
	\end{tabular}
	\caption{Dataset contents}
	\label{tbl-size}
\end{table}

As shown in Table \ref{tbl-size}, the dataset contains the metadata of 250,163 Scratch projects, and the code of 233,491 of those that are non-empty.
The projects were created by 109,960 different users.
Most of the users, as shown in Table~\ref{tbl-stats}, have created a single project.
However, 40,000 users have created more than one and up to 868 projects in the dataset.
Not all the projects were successfully analyzed with Dr. Scratch. This is due to problems with non-official extension blocks and with some non-ASCII characters in sprites names.

>>>>

A number of works in the computing education research field attempt to assess the programming skills that novice programmers develop in the Scratch environment.
Some utilize program data collected during specific programming courses (e.g.,~\cite{meerbaum-salant_learning_2010, wilson_evaluation_2012, Maloney_2008}), while others utilize the dataset made available by the Lifelong Kindergarten Group at the MIT Media Lab~\cite{2017_scratch_dataset}, which contains data for Scratch projects created until 2012 (e.g.,~\cite{fields_2014, yang_2015, Dasgupta_2016}).
In addition to identifying indications of learning of programming concepts, static analysis of Scratch programs has also been performed for identifying code smells and bad programming practices ~\cite{Meerbaum_habits_2011, Aivaloglou_2016}, and automated quality assessment tools have been proposed (e.g., Hairball~\cite{boe_hairball:_2013} and Dr. Scratch~\cite{moreno_automatic_2014}).

While Scratch is receiving increasing interest as an introductory programming language, there is no recent dataset of Scratch programs available to the research community.
The one made available from the MIT Media Lab concerns projects created using the previous, initial version of the Scratch application, before the introduction of the current Scratch version (Scratch 2) and of the web programming interface in 2013.
It is since then that the popularity of Scratch started to increase\footnote{Monthly activity trends can be found at \url{https://scratch.mit.edu/statistics/}}.

@@

\subsection{Data collection}
To collect the data from the web interface of the Scratch project repository, we built a scraping program.
The web scraping program starts by reading the Scratch projects page\footnote{\label{scratchpublic}\url{https://scratch.mit.edu/explore/projects/all/}} and thus obtains the project identifiers of projects that were most recently shared.
Subsequently, it retrieves a JSON file for each of the listed projects\footnote{For a given project id $x$, the program's JSON representation can be obtained via \url{https://cdn.projects.scratch.mit.edu/internalapi/project/x/get}}.

We ran the scraper on March 2nd 2016 for 24 hours and, during that time, it obtained the JSON files for \nPrograms~projects. Out of those, we failed to parse and further analyze 2,367 projects due to technical difficulties with the JSON files.
The scraper and all project files are available\footnotemark[\ref{dataseturl}].

Once we obtained the Scratch projects, we parsed the JSON files according to the specification of the format\footnote{\url{http://wiki.scratch.mit.edu/wiki/Scratch_File_Format_(2.0)}}.
This resulted in a list of used blocks per project, within the sprites and the stage of the project.
We also cross referenced all blocks with the Scratch wiki to determine their shapes and category.
For example, \texttt{When Green Flag Clicked} is a \emph{Hat block} from the \emph{Events category}.
We included blocks from Scratch extensions, such as the LEGO WeDo extensions, that were found in the dataset.

\subsection{Calculation of programming mastery scores}
%\fenia{Description of how the mastery metrics for       
%	[Abstraction]
%	,[Parallelism]
%	,[Logic]
%	,[Synchronization]
%	,[FlowControl]
%	,[UserInteractivity]
%	,[DataRepresentation]
%	,[Mastery]
%	,[Clones]
%	,[CustomBlocks]
%	,[InstancesSprites]
%	were calculated}
We analyzed the projects with Dr. Scratch, a tool that statically inspects Scratch projects' source code to assign scores on seven dimensions of computational thinking: abstraction and problem decomposition, logical thinking, synchronization, parallelism, algorithmic notions of flow control, user interactivity and data representation. These dimensions are given a value from 0 to 3. The mastery score is the aggregated punctuation of the seven dimensions, and therefore ranges from 0 to 21.

@@

Out of the \nPrograms~projects, 231,050 were successfully analyzed with Dr. Scratch. This was due to problems with non-official extension blocks and with some non-ASCII characters in sprites names.

@@

\begin{table}[]
	\centering
	\begin{tabular}{llp{5.4cm}}
		\textbf{Table}& \textbf{Key} & \textbf{Attribute(Description)}\\
		\hline
		\textbf{\texttt{project}} & PK & \texttt{p\_id} (Scratch project ID)\\
		&  & 							\texttt{project-name} (name given to project)\\
		&  & 							\texttt{username} (author's Scratch 
		username)\\
		&  & 							\texttt{total-views} (project views number)\\
		&  & 							\texttt{total-remixes} (project remixes number)\\
		&  & 							\texttt{total-favorites} (total users favoriting)\\
		&  & 							\texttt{total-loves} (users `loving' the project)\\
		& & 							\texttt{is-remix} (calculated column, true if project is a remix of another one)\\
		\hline
		\textbf{\texttt{script}} & PK & \texttt{script\_id} (auto increment)\\
		& FK & 							\texttt{project\_id} (\texttt{\scriptsize{*\ldots1 project:p\_id}})\\
		& & 							\texttt{sprite-type} ([$sprite \vert stage \vert procDef$])\\
		& & 							\texttt{sprite-name} (name of sprite the script is in)\\
		& & 							\texttt{script-rank} (project-level ranking of script)\\
		& & 							\texttt{coordinates} (x-y location of script in Scratch editor)\\
		& & 							\texttt{total-blocks} (calculated column, number of blocks comprising the script)\\
		\hline
		\textbf{\texttt{procedure}} & PK,FK & \texttt{script\_id} (\texttt{\scriptsize{1\ldots1 script:script\_id}})\\
		& & 							\texttt{proc-name} (name given to custom block)\\
		& & 							\texttt{total-args} (number of input arguments)\\
		\hline
		\textbf{\texttt{block}} & PK,FK  & \texttt{script\_id} (\texttt{\scriptsize{*\ldots1 script:script\_id}})\\
		& PK & \texttt{block-rank} (script-level ranking of block)\\
		&  & \texttt{block-type} (\texttt{\scriptsize{*\ldots1 blockType:b-type}})\\
		&  & \texttt{parameter-1} (value of 1st input parameter)\\
		&  & ... \\
		&  & \texttt{parameter-24} (value 24th input parameter)\\
		\hline
		\textbf{\texttt{blockType}} & PK & \texttt{b-type} (Scratch name of predefined block)\\
		&  & \texttt{category} (Scratch block category)\\
		&  & \texttt{shape} (One of 8 block shapes in Scratch)\\
		&  & \texttt{is-input} (true if block receives user input)\\
		\hline
		\textbf{\texttt{remix}} & PK  & \texttt{remix\_id} (\texttt{\scriptsize{project\_id}} of created project)\\
		&  & \texttt{from-project\_id} (\texttt{\scriptsize{project\_id}} of original project)\\
		\hline
		\textbf{\texttt{grades}} & PK,FK & \texttt{project\_id} (\texttt{\scriptsize{1\ldots1 project:p\_id}})\\
% Original
%		& & \texttt{Abstraction} (Score for abstraction and problem decomposition)\\
%		& & \texttt{Parallelism} (Score for parallelism)\\
%		& & \texttt{Logic} (Score for logical thinking)\\
%		& & \texttt{Synchronization} (Score for synchronization)\\
%		& & \texttt{FlowControl} (Score for algorithmic notions of flow control)\\
%		& & \texttt{UserInteractivity} (Score for user interactivity)\\
%		& & \texttt{DataRepresentation} (Score for data representation)\\

		& & \texttt{Abstraction} (Score)\\
		& & \texttt{Parallelism} (Score)\\
		& & \texttt{Logic} (Score)\\
		& & \texttt{Synchronization} (Score)\\
		& & \texttt{FlowControl} (Score)\\
		& & \texttt{UserInteractivity} (Score)\\
		& & \texttt{DataRepresentation} (Score)\\
		& & \texttt{Mastery} (Dr. Scratch total mastery score)\\
		& & \texttt{Clones} (number of software clones)\\
		& & \texttt{CustomBlocks} (number of user defined blocks)\\
		& & \texttt{InstancesSprites} (true if project uses Scratch clones)\\
		\hline
	\end{tabular}
	\caption{Database schema: Tables and Attributes}
	\label{tbl-dbschema}
\end{table}

@@

The projects' data are stored in a relational database. Each \textbf{\texttt{project}} is identified by its Scratch project ID, stored in field \texttt{p\_id}, while its author is identified by the \texttt{username}\footnote{\label{fn-authorpage}Each project and author page can be accessed in the Scratch web interface at \url{https://scratch.mit.edu/projects/<p_id>} and \url{https://scratch.mit.edu/users/<username>} respectively}.
If a project is a remix of another one, the original project can be found in the \textbf{\texttt{remix}} table.
Table~\textbf{\texttt{grades}} stores the Dr. Scratch results for the programming mastery metrics per project.

The schema of the database, outlined in Table~\ref{tbl-dbschema}, reflects the structure of the Scratch programs.
Projects contain sprites, which are entities with their own associated code.
The code is organized into \textbf{\texttt{script}}s, i.e., groups of Scratch code blocks, each script belonging to a sprite named \texttt{sprite-name}.
In the example in Figure~\ref{fig-scratchExample} there are 3 scripts, one initiated when the green flag is clicked, one then the sprite is clicked, and the `backflip' custom block definition.
Those custom blocks are the equivalent of procedure definitions, with their names and arguments stored in the \textbf{\texttt{procedure}} table.

@@

The coding elements in the Scratch environment are the \textbf{\texttt{block}}s, which can receive input parameters.
In the example in the figure, block `go to x:0 y:0' receives as input two numerical parameters.
Custom clock invocations are stored in this table just like regular blocks.
To cater for the varying number of input parameters that those blocks can have, this table can store up to the maximum number of input parameters found in the dataset, which is 24.
The ordering of the blocks in the scripts can be retrieved using the \texttt{block-rank} field, which represents a depth-first ranking of the blocks.
In the script of the `backflip' custom block definition in the example, the ranking of the blocks would be `define backflip', `set counter to', `+', `counter', followed by the rest of the blocks.
The clocks can be of different shapes and categories.
In table \textbf{\texttt{blockType}} we have stored the encoding of all types of Scratch blocks and of the blocks from Scratch extensions.

@@

\begin{table}[]
	\centering
	\begin{tabular}{lr}
		\hline
		Projects & 250,163 \\
		Non-empty projects & 233,491 \\
		Projects analyzed by Dr. Scratch & 231,050 \\
		Remixes of other projects & 12,167 \\
		Usernames & 109,960 \\
		Scripts & 4,049,356 \\
		Blocks & 36,085,654 \\
		Custom block definitions & 204,018 \\
		\hline
	\end{tabular}
	\caption{Dataset contents}
	\label{tbl-size}
\end{table}

As shown in Table \ref{tbl-size}, the dataset contains the metadata of 250,163 Scratch projects, and the code of 233,491 of those that are non-empty.
The projects were created by 109,960 different users.
Most of the users, as shown in Table~\ref{tbl-stats}, have created a single project.
However, 40,000 users have created more than one and up to 868 projects in the dataset.
